Sperm should evolve to make female meiosis fair.

### 2014-07-11 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
begin revisions

%%!TEX TS-program = latex
\documentclass[12pt,letterpaper]{article}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\bibliographystyle{genetics}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}  
	\newcommand{\gc}[1]{{ \color{red} #1}}
	\newcommand{\yb}[1]{{ \color{blue} #1}}

%Why do animals allow male control of female meiosis?
%Why does female meiosis allow the opportunity for exploitation by selfish sperm?
%sperm and egg genomes both favour the peaceful resolution of female meiotic drive
%strange bed fellows

%\title{Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?}
\title{
%Co-operation between the sexes trumps selfish control of female  meiotic drive. 
%Or 
Sperm should evolve to make female meiosis fair.
%Sperm do not evolve to collaborate in female meiotic drive. 
%Or Why does female meiosis allow the opportunity for exploitation by selfish sperm? 
}

\author{Yaniv Brandvain$^1$ \\ email: ybrandvain@gmail.com  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
$^1$Department of Plant Biology, \\ University of Minnesota --€“ Twin
Cities. St. Paul MN, 55108 \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ University of California -- Davis. Davis, CA, 95616
\end{center}

%FIgures 
{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
%Because only one of four products of female meiosis is transmitted to offspring, 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
%(meiotic drive).  
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. % drive the rapid evolution of suppression mechanisms, and
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? %and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. %%felt this
                                %%might be an important addition as
                                %%they will spread if no cost to organism
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        %, and variation in stage of meiosis that
        %fertilization occurs at 
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis. %prevent meiotic drive.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals  
	female meiosis is complete upon ovulation (that is, fertilization is not required for female meiosis \citep[e.g. Table 1 in 
][]{Masui_book}).

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster -- a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. %, focussing on models in which `self-promoting' alleles in sperm facilitate drive of like genotypes during gametogenesis in heterozygous females. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. %The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}.
	%YB - MOVED THIS TO THE APPENDIX
          To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm
        carrying that allele (see Figure \ref{Eggsperm}A and Model 2
        in the Appendix and Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude our single locus results with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we first briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming that the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$, solid black line in Figure \ref{Eggsperm}B) a driver cannot fix and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

\begin{figure}
\begin{center}
\includegraphics[width = 1 \textwidth]{Figures/selfprom.eps} \end{center}
\caption{{\bf{A.}} A visual depiction of our model of `self-promoting' driver. 
	Transmission probabilities for alleles through female
 	 meiosis depend on sperm genotype. 
	 The non-driving $A$-allele, and self-promoting $B$-allele
	 are represented by unfilled and filled circles, respectively. 
	 {\bf{B.}} Evolution of a self-promoter and standard driver. 
	 	Assuming that the fitnesses of drive homozygotes and 
		heterozygotes are $1-s$ and $1$, respectively. 
	 	\emph{Main figure:} Boundary conditions for the invasion 
		and fixation of self-promoting and standard meiotic drivers, 
		with drive coefficient, $d$. Colored regions depict exact results, 
		while lines  represent analytical approximations. 
	 \emph{B1:}  Trajectories of sperm-dependent female drive
		 %alleles that can invade the population, 
		 each allele has $s =0.1$ against the
		 homozygotes.  The drive coefficient is denoted by color. 
 }   
\label{Eggsperm}
\end{figure}

@@

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggests a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3 in the Appendix) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
	% fixes when $s$ is less than $d_{hom} - 1/2$.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure \ref{Effect_of_dominance}, and Equation \ref{polymale}).  
%Presumably this zone of polymorphism comes about because positive frequency 
%	dependence diminishes with the dominance of the male effect driver. 
%As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver (assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
%Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
%Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt its general applicability. 
%Thus we regard this case as interesting but not very general.

@@

%\subsection*{ sperm-egg haplotype balanced drive systems}

@@

%%%sperm-acting facilitator
When the modifier of drive arises on the drive background (i.e. in coupling phase),
	is tightly linked to the driver, and enhances drive 
	we label  this non-recombining drive/modifier haplotype as the $B^+$ allele.  
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$
        heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}A2, and
        Model 4 in the Appendix and Supplementary Worksheet). 
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, 
	the novel $B^{+}$ haplotype
	arises when the ancestral driver 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population (Appendix Model 4, eqn. \eqref{coupling}, and Mathematical Supplement).

\begin{figure}
\begin{center}
	\includegraphics[width = 0.7
          \textwidth]{Figures/twolocus.eps} \end{center}
\caption{Models of a sperm-acting drive modifier tightly linked to a meiotic driver. 
{\bf (A) } 
Sperm carrying the derived allele at the modifier locus (filled squares) alters transmission  
	at the driving allele (filled circles) during female meiosis.
        Alleles at these two tightly linked loci form three haplotypes (top of A). 	
	{\bf{A1)}} In the standard model of drive there is no
        variation at the modifier, 
	and the driver is transmitted to the egg with probability $d_0$. 
	{\bf{A2)}} The modifier allele increases the transmission of the drive allele ($d_1 > d_0$), and due to their shared genetic background, also increases its drive. 
	{\bf{A3 \& A4)}} The sperm-acting modifier acts to decrease
        drive  ($d_1=1/2$ in A3 \& A4, or more generally, $d_1 < d_0$) and arises on the same or opposite
        background from the driver (A3 \& A4 respectively). 
%		In {\bf{A3}} the drive suppressor exists on the drive background, forming the $B^-$ haplotype. 
%		In {\bf{A4}} the drive suppressor exists on the non-drive background, forming the $A^-$ haplotype. 		
{\bf{(B)}} Invasion of a sperm-acting drive suppressor linked to a driver. 
	After the driver ($B$ haplotype) reaches drive selection equilibrium, we introduce a sperm acting drive modifier. 
	We assume full drive ($D_0 =1$), a recessive lethal fitness cost to drive ($h_s=0$, $s=1$) and that the sperm-acting modifier results in a fair meiosis. 
{\bf {B1)}} The $B^-$ allele replaces the ancestral drive haplotype, but segregates at a lower equilibrium frequency.
{\bf {B2)}} The $A^-$ allele replaces the ancestral non-driving haplotype, and in this case, removes the driver from the population. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
%	that is sperm-acting alleles that act to restore  fairness of female meiosis, can. %If these sperm-acting drive suppressors arise on
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE.
This allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.

%More surprisingly, sperm-acting suppressors of female drive can spread
%	even when tightly linked to the original drive system (Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
%The $B^{-}$ haplotype (representing the tightly linked drive/drive suppressor complex) displaces the original driver 
%	by driving when not costly, but avoiding low fitness homozygotes by disrupting drive when transmitted to sperm. %  ($BB^-$ and $B^-B^-$). 
%Despite displacing the ancestral driver (the $B$ haplotype), the equilibrium frequency of the $B^-$ haplotype is often lower than that of $B$ \yb{(point to supp?)}.

%\gc{Do we mention the In and Om locus here or in the discussion as a
%	 potential e.g. Do we mention the fact that the frequency of these
%	 newly invade sperm dodging alleles can be lower than the original system?}\\

@@

\begin{figure}

>>>>

%%!TEX TS-program = latex
\documentclass[12pt,letterpaper]{article}
\bibliographystyle{genetics}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}  
	\newcommand{\gc}[1]{{ \color{red} #1}}
	\newcommand{\yb}[1]{{ \color{blue} #1}}

\title{  Sperm should evolve to make female meiosis fair.  }

\author{Yaniv Brandvain$^1$ \and Graham Coop$^2$ }

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\begin{document}
\maketitle
\begin{center}
$^1$Department of Plant Biology, \\ University of Minnesota --€“ Twin
Cities. St. Paul MN, 55108 \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ University of California -- Davis. Davis, CA, 95616
\end{center}

\newpage 
\flushleft{{\bf{Running title:}} Sperm prevent female meiotic drive} \\
{\bf{Keywords:}} Meiosis; Meiotic drive; Fertilization; Genomic conflict \\
{\bf{Corresponding author:}} Yaniv Brandvain. \emph{Address:} Department of Plant Biology, University of Minnesota --€“ Twin
Cities. St. Paul MN, 55108. \emph{Phone:} 612-624-4375 \emph{Email:} ybrandvain@gmail.com

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems.  
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

%It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
%The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals  
%	female meiosis is complete upon ovulation (that is, fertilization is not required for female meiosis \citep[e.g. Table 1 in 
%][]{Masui_book}).

%It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
%In \emph{C. elegans}, premature deployment of the aster -- a vital component of mitotic
%machinery provided by the sperm, disrupts MII meiotic segregation
%in the egg, leading to a triploid zygote \citep{McNally2012}. 
%Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
% Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
% transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
% While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
%52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
\yb{While, we do not claim that female meiotic arrest evolved specifically to recruit helpful sperm to defend against meiotic drive, the delayed completion of female meiotic drive will not in fact encourage sperm to assist renegade drivers.}
%Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
%	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. 
 To depict a case of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm
        carrying that allele (see Figure \ref{Eggsperm}A and Model 2
        in the Appendix and File S1 ). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude our single locus results with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we first briefly present the standard drive model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming that the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$, solid black line in Figure \ref{Eggsperm}B) a driver cannot fix and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

\begin{figure}
\begin{center}
\includegraphics[width = 1 \textwidth]{../Figures/selfprom.eps} \end{center}
\caption{{\bf{A.}} A visual depiction of our model of `self-promoting' driver. 
	Transmission probabilities for alleles through female
 	 meiosis depend on sperm genotype. 
	 The non-driving $A$-allele, and self-promoting $B$-allele
	 are represented by unfilled and filled circles, respectively. 
	 {\bf{B.}} Evolution of a self-promoter and standard driver. 
	 	Assuming that the fitnesses of drive homozygotes and 
		heterozygotes are $1-s$ and $1$, respectively. 
	 	\emph{Main figure:} Boundary conditions for the invasion 
		and fixation of self-promoting and standard meiotic drivers, 
		with drive coefficient, $d$. Colored regions depict exact results, 
		while lines  represent analytical approximations. 
	 \emph{B1:}  Trajectories of sperm-dependent female drive
		 %alleles that can invade the population, 
		 each allele has $s =0.1$ against the
		 homozygotes.  The drive coefficient is denoted by color. 
 }   
\label{Eggsperm}
\end{figure}

@@

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggests a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3 in the Appendix) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
	% fixes when $s$ is less than $d_{hom} - 1/2$.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure \ref{Effect_of_dominance}, and Equation \ref{polymale}).  
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt its general applicability.

@@

@@

%%%sperm-acting facilitator
When the modifier of drive arises on the drive background (i.e. in coupling phase),
	is tightly linked to the driver, and enhances drive 
	we label  this non-recombining drive/modifier haplotype as the $B^+$ allele.  
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$
        heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}A2, and
        Model 4 in the Appendix and File S1 ). 
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, 
	the novel $B^{+}$ haplotype
	arises when the ancestral driver 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population (Appendix Model 4, eqn. \eqref{coupling}, and File S2  ).

\begin{figure}
\begin{center}
	\includegraphics[width = 0.7 \textwidth]{../Figures/twolocus.eps} 
          \end{center}
\caption{Models of a sperm-acting drive modifier tightly linked to a meiotic driver. 
{\bf (A) } 
Sperm carrying the derived allele at the modifier locus (filled squares) alters transmission  
	at the driving allele (filled circles) during female meiosis.
        Alleles at these two tightly linked loci form three haplotypes (top of A). 	
	{\bf{A1)}} In the standard model of drive there is no
        variation at the modifier, 
	and the driver is transmitted to the egg with probability $d_0$. 
	{\bf{A2)}} The modifier allele increases the transmission of the drive allele ($d_1 > d_0$), and due to their shared genetic background, also increases its drive. 
	{\bf{A3 \& A4)}} The sperm-acting modifier acts to decrease
        drive  ($d_1=1/2$ in A3 \& A4, or more generally, $d_1 < d_0$) and arises on the same or opposite
        background from the driver (A3 \& A4 respectively). 
{\bf{(B)}} Invasion of a sperm-acting drive suppressor linked to a driver. 
	After the driver ($B$ haplotype) reaches drive selection equilibrium, we introduce a sperm acting drive modifier. 
	We assume full drive ($D_0 =1$), a recessive lethal fitness cost to drive ($h_s=0$, $s=1$) and that the sperm-acting modifier results in a fair meiosis. 
{\bf {B1)}} The $B^-$ allele replaces the ancestral drive haplotype, but segregates at a lower equilibrium frequency.
{\bf {B2)}} The $A^-$ allele replaces the ancestral non-driving haplotype, and in this case, removes the driver from the population. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE.
This allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.

@@

{\bf{File S1:}} Transmission rules. We detail the frequency of offspring genotypes produced by each possible mating for Models 1-6 in the tabs of this Excel spreadsheet.

{\bf{Files S2A \& S2B:}} A Mathematica file (FileS2A) and a PDF of this file (File S2B) in which we derive analytical results for models 1-6.
